What does the crowd believe? A hierarchical approach to estimating subjective beliefs from empirical data

### 2016-01-31 text/01_CogSci_abstract/paper.tex
edits to motivation

\definecolor{Green}{RGB}{10,200,100}
\newcommand{\ndg}[1]{\textcolor{Green}{[ndg: #1]}}  
\definecolor{Blue}{RGB}{10,100,200}
\newcommand{\jd}[1]{\textcolor{Blue}{[jd: #1]}} 
\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[mf: #1]}}

@@

We cannot look into a person's head. But in making sense of observed behavior we readily
ascribe beliefs and desires to fellow agents. This happens intuitively, in folk psychology, but
also in science. Scientific ascription of latent mental states plays an important role in many
explanations of higher-order cognition: decision making, planning, reasoning, and language
use. It is therefore vital to have methods for validating explanatory mental states
ascriptions.

@@

In sum, there is tradeoff between simplicity of paradigms and their informativity. Ideally, we
would like to have an experimental method for measuring subjective beliefs that (i) provides
sufficient and reliable-enough information about subjective beliefs to derive testable
predictions from cognitive models that rely on such information, (ii) is easy to understand by
 participants, (iii) is easy to implement by researchers whose focus and expertise is not
belief-elicitation as such, and that (iv) does not rely on sophisticated methods of extracting
the relevant information from the observed data. Moreover, the ideal method would be (v)
flexible enough to allow inferred subjective belief distributions beyond standard parameterized
distributions and (vi) would provide us with a reasonable approximation of a population-level
belief, i.e., an estimate of the central tendency or average of subjective beliefs in a given
population. The latter would warrant using one set of participants for measuring prior beliefs and
another for whatever other task is of interest.

A simple technique that seems to fit this bill is the \emph{binned histogram} task that has
recently been used with apparent success
\citep{KaoWu2014:Nonliteral-Unde,SchollerFranke2015:Semantic-values}. \mf{someone please add
  references: where did you use binned sliders?} Participants adjust sliders in order to
express their (relative) subjective beliefs about how likely it is that the value of an
uncertain contiguous quantity lies in some interval of possible values (a bin). E.g.,
participants read that John bought a watch.  To report their beliefs about the watch's price,
they adjust sliders whose endpoints are labelled as ``extremely likely'' and ``extremely
unlikely.'' Each slider corresponds to one out of 15 bins that partition the range of plausible
values (established by a pre-test). E.g., for the price of a watch bins could be ``\$0-\$50'',
``\$50-\$100'', etc. up to ``\$700-\$750'' and ``more than \$750''. Each participant's slider
ratings are normalized within each item, and the results are averaged across participants to obtain
an estimate of the population's average belief about each item. These averaged normalized
slider ratings per bin (henceforth: mean slider ratings) look intuitively plausible and yield
good results if fed into cognitive models that predict behavior in language or reasoning tasks,
for instance.

Explanatory success notwithstanding, the question remains whether mean slider ratings measure
what we would like them to. Are these good approximations of a central tendency of participants'
individual beliefs? Are these measures consistent with participants' behavior in other tasks, such
as \emph{give-a-number}, as well? To address this issue we collected data in a within-subject
from three task types: (i) \emph{binned histograms}, (ii) \emph{give-a-number} and (iii)
\emph{paired comparisons}, i.e., a direct comparison of two bins from the \emph{binned
  histograms} task. Using a Bayesian hierarchical model, we used the data from all three
dependent measures to infer likely latent subjective beliefs, where each subjective belief is,
intuitively, a noisy perturbation of a population-level belief. We find that mean \emph{a
  posteriori} population-level beliefs are approximated well by mean slider ratings, suggesting
that the latter may indeed be a plausible, yet cheap measure of what the crowd believes.
Moreover, the model is able to capture participants' behavior in all three task types reasonably
well, suggesting that what mean slider ratings measure is what we think it is: a population-level
central tendency of latent subjective beliefs.

>>>>

\definecolor{Green}{RGB}{10,200,100}
\newcommand{\ndg}[1]{\textcolor{Green}{[ndg: #1]}}  
\definecolor{Blue}{RGB}{10,100,200}
\newcommand{\jd}[1]{\textcolor{Blue}{[jd: #1]}} 
\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[mf: #1]}} 
\definecolor{DarkOrange}{RGB}{255,100,50}
\newcommand{\mht}[1]{\textcolor{DarkOrange}{[mht: #1]}}

@@

We cannot look into a person's head. But in making sense of observed behavior we readily
ascribe beliefs and desires to fellow agents. This happens intuitively, in folk psychology, but
also in science. Scientific ascription of latent mental states plays an important role in many
explanations of higher-order cognition: decision making, planning, reasoning, and language
use. It is therefore vital to have methods for validating explanatory mental state
ascriptions.

@@

In sum, there is tradeoff between simplicity of paradigms and their information content. Ideally, we
would like to have an experimental method for measuring subjective beliefs that (i) provides
sufficient and reliable-enough information about subjective beliefs to derive testable
predictions from cognitive models that rely on such information, (ii) is easy to understand by
 participants, (iii) is easy to implement by researchers whose focus and expertise is not
belief-elicitation as such, and that (iv) does not rely on sophisticated methods of extracting
the relevant information from the observed data. Moreover, the ideal method would be (v)
flexible enough to allow inferred subjective belief distributions beyond standard parameterized
distributions and (vi) would provide us with a reasonable approximation of a population-level
belief, i.e., an estimate of the central tendency or average of subjective beliefs in a given
population. The latter would warrant using one set of participants for measuring prior beliefs and
another for whatever other task is of interest.

A simple technique that seems to fit this bill is the \emph{binned histogram} task that has
recently been used with apparent success
\citep{KaoWu2014:Nonliteral-Unde,SchollerFranke2015:Semantic-values}. \mf{someone please add
  references: where did you use binned sliders?} \mht{I used ``categorical'' sliders for discrete priors in my syllogistic reasoning / belief bias stuff; and so did justine in her metaphor work...}Participants adjust sliders in order to
express their (relative) subjective beliefs about how likely it is that the value of an
uncertain contiguous quantity lies in some interval of possible values (a bin). E.g.,
participants read that John bought a watch.  To report their beliefs about the watch's price,
they adjust sliders whose endpoints are labelled as ``extremely likely'' and ``extremely
unlikely.'' Each slider corresponds to one out of 15 bins that partition the range of plausible
values (established by a pre-test). E.g., for the price of a watch bins could be ``\$0-\$50'',
``\$50-\$100'', etc. up to ``\$700-\$750'' and ``more than \$750''. Each participant's slider
ratings are normalized within each item, and the results are averaged across participants to obtain
an estimate of the population's average belief about each item. These averaged normalized
slider ratings per bin (henceforth: mean slider ratings) look intuitively plausible and yield
good results if fed into cognitive models that predict behavior in language or reasoning tasks,
for instance.

Explanatory success notwithstanding, the question remains whether mean slider ratings measure
what we would like them to. Are these good approximations of a central tendency of participants'
individual beliefs? Are these measures consistent with participants' behavior in other tasks, such
as \emph{give-a-number}, as well? To address this issue we collected data in a within-subject paradigm 
from three task types: (i) \emph{binned histograms}, (ii) \emph{give-a-number} and (iii)
\emph{paired comparisons}, i.e., a direct comparison of two bins from the \emph{binned
  histograms} task. Using a Bayesian hierarchical model, we used the data from all three
dependent measures to infer likely latent subjective beliefs, where each subjective belief is,
intuitively, a noisy perturbation of a population-level belief. We find that mean \emph{a
  posteriori} population-level beliefs are approximated well by mean slider ratings, suggesting
that the latter may indeed be a plausible, yet cheap measure of what the crowd believes.
Moreover, the model is able to capture participants' behavior in all three task types reasonably
well, suggesting that what mean slider ratings measure is what we think it is: a population-level
central tendency of latent subjective beliefs.
